Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-19 paper.tex
Reformulate naive algorithm in terms of matches

Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some (though not all) synthetic benchmarks.
The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement}.
Mainline Aesop already uses the new implementation by default (TODO true?).

@@

\jcom{Formulate this in terms of matches.}
Prior to this work, Aesop used what we call naive forward reasoning~\cite{Aesop}.
This approach treats forward rules essentially like any other Aesop rule.
When Aesop selects a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we first unify the last premise $Aₘ$ with each hypothesis $h : C$ such that $C$ and $Aₘ$ likely unify (as determined by a discrimination tree index).
If successful, this results in a substitution $σ₁$ containing $xₘ$ and any variables occurring in $Aₘ$.
We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$ (skipping premises $A_{m-i}$ if they are already determined by $σᵢ$).
If a premise $A_{m-i}[σᵢ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
These may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
If this process yields at least one full application $r~a₁ \dots aₘ$, we can add a new hypothesis of type $T = B[x₁ ↦ a₁, \dots, xₘ ↦ aₘ]$, meaning we generate the subgoal $Γ, h : T$ (unless $Γ$ already contains a hypothesis of type $T$).

This approach is efficient when the last premise of a rule is not present in the context since the rule is then excluded by indexing and never run.
If the last premise is present, however, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically~\cite{TammetSubsumption}).
Furthermore, the whole process is repeated when we proceed to another goal with a similar context.
In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be substantially similar, if not identical, to the current context, but these similarities are not exploited by the naive algorithm.

@@

% TODO pre- and postcondition?
\begin{algorithm}
  \Input{A partial match $m$ of $r$ at level $i$ with substitution $σ$}
  \Output{A set of complete matches of $r$}
  \BlankLine
  \Procedure{$\mathrm{addMatch}(m)$}{
    \eIf{$i$ is complete}{
      \Return{$\{m\}$}
    }{
      \For{$x$ \KwIn $\vars(i + 1)$}{
        Insert $m$ into the set $M_{x}(i, σ(x))$
      }
      \KwLet $\mathit{hs} ≔ \bigcap_{x \in \vars(i + 1)} H_{x}(i + 1, σ(x))$ \\
      $\mathit{completeMs} \gets ∅$ \\
      \For{$h$ \KwIn $\mathit{hs}$}{
        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m \mathbin{++} [h])$
      }
      \Return{$\mathit{completeMs}$}
    }
  }
  \BlankLine % without these, the caption is not offset properly
  \BlankLine
  \caption{Add a match to the rule state for a rule $r$}%
  \label{alg:addmatch}
\end{algorithm}

@@

There are unfortunately no standard benchmarks for Aesop-like tactics, and since Aesop is not expected to solve most goals without an extensive, manually curated set of rules, benchmarks for push-button theorem provers such as TPTP~\cite{TPTP} are not very meaningful.
Additionally, forward rules are currently little used by Aesop's main client, the Mathlib mathematics library~\cite{Mathlib}.
We therefore only evaluate our technique against the previous, naive Aesop implementation on several synthetic benchmarks that demonstrate the performance characteristics of incremental forward reasoning.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of hypotheses},
        ylabel={Time in ms},
        xmin=1, xmax=32,
        ymin=4, ymax=2^16,
        %xtick={1,2,4,8,16},
        %ytick={2^3,2^5,2^7,2^9,2^11,2^13},
        xmode=log,
        ymode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=}
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111) (32, 45220.919667)  
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681) (32, 1175.888750)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417) (32, 392049.849708) 
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680) (32, 4388.917500) 
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Transitivity benchmark}%
    \label{fig:trans}
  \end{subfigure}%
  \begin{subfigure}[b]{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of rules},
        %ylabel={Time in ms},
        xmin=1, xmax=32,
        ymin=4, ymax=2^16,
        ymode=log,xmode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 10.399958) (2, 16.426541) (4, 43.461403) (8, 205.595541) (16, 1238.090889) (32, 8927.857819)
          };
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 11.231875) (2, 12.879500) (4, 16.851930) (8, 24.666485) (16, 42.054055) (32, 76.369778)
          };
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 19.405222) (2, 33.487472) (4, 80.107597) (8, 284.954791) (16, 1536.089624) (32, 10721.729861)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.892069) (2, 40.031930) (4, 69.446902) (8, 130.640986) (16, 260.653917) (32, 481.827902)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Independence benchmark}%
    \label{fig:indep}
  \end{subfigure}
  \caption{Results of the transitivity (\subref{fig:trans}) and independence (\subref{fig:indep}) benchmarks for $a = 0$ (\protect\showpgfsquare) and $a = 100$ (\protect\showpgfcircle).}%
  \label{fig:benchmark}
  \end{figure}

@@

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Depth},
        ylabel={Time in ms},
        xmin=1, xmax=5,
        ymin=8, ymax=2^10,
        xtick={1,2,3,4,5},
        ymode=log,
        log basis y=2,
        ymajorgrids=true,
        grid style=dashed
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 35.466292) (2, 40.918638) (3, 46.936125) (4, 54.933124) (5, 61.269861)
          };
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 135.967736) (2, 127.324444) (3, 136.891250) (4, 118.720444) (5, 116.293222)
          };
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 107.978069) (2, 115.900527) (3, 122.454958) (4, 137.732764) (5, 144.037681)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 619.932611) (2, 601.456805) (3, 600.926110) (4, 589.144389) (5, 595.091180)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Without precompilation}%
    \label{fig:erase-noprecomp}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Depth},
        xmin=1, xmax=5,
        ymin=8, ymax=1024,
        xtick={1,2,3,4,5},
        ymode=log,
        log basis y=2,
        yticklabels={},
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 13.728805) (2, 17.909194) (3, 23.324666) (4, 28.030819) (5, 33.279611)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 13.841305) (2, 12.801791) (3, 13.358389) (4, 13.200611) (5, 13.251902)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 81.546486) (2, 85.345444) (3, 90.532347) (4, 98.247764) (5, 105.423875)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.539097) (2, 24.432485) (3, 25.629666) (4, 23.935250) (5, 23.911139)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{With precompilation}%
    \label{fig:erase-precomp}
  \end{subfigure}
  \caption{Results of the depth benchmark with precompilation disabled (\subref{fig:erase-noprecomp}) or enabled (\subref{fig:erase-precomp}) and with $a = 0$ (\protect\showpgfsquare) or $a = 100$ (\protect\showpgfcircle).}%
  \label{fig:erase}
\end{figure}

>>>>

Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some synthetic benchmarks.
The implementation, benchmarks and evaluation data are available in a supplement to this paper~\cite{supplement} and the new implementation is already available in mainline Aesop.

@@

Prior to this work, Aesop used what we call \emph{naive forward reasoning}~\cite{Aesop}.
This approach treats forward rules essentially like any other Aesop rule.
Hence, when Aesop selects a forward rule $r$ with $m$ slots of types $A₁, \dots, Aₘ$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we look for all complete matches for $r$ in $Γ$.

To that end, we first unify the type of the last slot, $Aₘ$, with each hypothesis $h : B$ such that $B$ and $Aₘ$ likely unify (as determined by a discrimination tree index).
If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₘ$.
We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$.
If a premise $A_{m-i}[σᵢ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
These may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis $h : B[σ]$, where $B$ is the conclusion of $r$.

This approach is efficient when the last slot of a rule is not present in the context since the rule is then excluded by indexing and never run.
Otherwise, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically~\cite{TammetSubsumption}).
Furthermore, the whole process is repeated when we proceed to another goal with a similar context.
In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be substantially similar, if not identical, to the current context, but these similarities are not exploited by the naive algorithm.

@@

\begin{algorithm}
  \Input{A partial match $m$ of $r$ at level $i$ with substitution $σ$}
  \Output{A set of complete matches of $r$}
  \BlankLine
  \Procedure{$\mathrm{addMatch}(m)$}{
    \eIf{$i$ is complete}{
      \Return{$\{m\}$}
    }{
      \For{$x$ \KwIn $\vars(i + 1)$}{
        Insert $m$ into the set $M_{x}(i, σ(x))$
      }
      \KwLet $\mathit{hs} ≔ \bigcap_{x \in \vars(i + 1)} H_{x}(i + 1, σ(x))$ \\
      $\mathit{completeMs} \gets ∅$ \\
      \For{$h$ \KwIn $\mathit{hs}$}{
        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m \mathbin{++} [h])$
      }
      \Return{$\mathit{completeMs}$}
    }
  }
  \BlankLine % without these, the caption is not offset properly
  \BlankLine
  \caption{Add a match to the rule state for a rule $r$}%
  \label{alg:addmatch}
\end{algorithm}

@@

There are unfortunately no standard benchmarks for Aesop-like tactics, and since Aesop is not expected to solve most goals without an extensive, manually curated set of rules, benchmarks for push-button theorem provers such as TPTP~\cite{TPTP} are not very meaningful.
Additionally, forward rules are little used by Aesop's main client, the Mathlib library~\cite{Mathlib}.
We therefore only evaluate our technique against the previous, naive Aesop implementation on several synthetic benchmarks that demonstrate the performance characteristics of incremental forward reasoning.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of hypotheses},
        ylabel={Time in ms},
        xmin=1, xmax=32,
        ymin=4, ymax=2^16,
        %xtick={1,2,4,8,16},
        %ytick={2^3,2^5,2^7,2^9,2^11,2^13},
        xmode=log,
        ymode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=}
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111) (32, 45220.919667)  
          };
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681) (32, 1175.888750)
          };
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417) (32, 392049.849708) 
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680) (32, 4388.917500) 
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Transitivity benchmark}%
    \label{fig:trans}
  \end{subfigure}%
  \begin{subfigure}[b]{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of rules},
        %ylabel={Time in ms},
        xmin=1, xmax=32,
        ymin=4, ymax=2^16,
        ymode=log,xmode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 10.399958) (2, 16.426541) (4, 43.461403) (8, 205.595541) (16, 1238.090889) (32, 8927.857819)
          };
          \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 11.231875) (2, 12.879500) (4, 16.851930) (8, 24.666485) (16, 42.054055) (32, 76.369778)
          };
          \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 19.405222) (2, 33.487472) (4, 80.107597) (8, 284.954791) (16, 1536.089624) (32, 10721.729861)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.892069) (2, 40.031930) (4, 69.446902) (8, 130.640986) (16, 260.653917) (32, 481.827902)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Independence benchmark}%
    \label{fig:indep}
  \end{subfigure}
  \caption{Results of the transitivity (\subref{fig:trans}) and independence (\subref{fig:indep}) benchmarks for $a = 0$ (\protect\showpgfsquare) and $a = 100$ (\protect\showpgfcircle).}%
  \label{fig:benchmark}
  \end{figure}

@@

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Depth},
        ylabel={Time in ms},
        xmin=1, xmax=5,
        ymin=8, ymax=2^10,
        xtick={1,2,3,4,5},
        ymode=log,
        log basis y=2,
        ymajorgrids=true,
        grid style=dashed
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 35.466292) (2, 40.918638) (3, 46.936125) (4, 54.933124) (5, 61.269861)
          };
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 135.967736) (2, 127.324444) (3, 136.891250) (4, 118.720444) (5, 116.293222)
          };
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 107.978069) (2, 115.900527) (3, 122.454958) (4, 137.732764) (5, 144.037681)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 619.932611) (2, 601.456805) (3, 600.926110) (4, 589.144389) (5, 595.091180)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Without precompilation}%
    \label{fig:erase-noprecomp}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Depth},
        xmin=1, xmax=5,
        ymin=8, ymax=1024,
        xtick={1,2,3,4,5},
        ymode=log,
        log basis y=2,
        %yticklabels={},
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 13.728805) (2, 17.909194) (3, 23.324666) (4, 28.030819) (5, 33.279611)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 13.841305) (2, 12.801791) (3, 13.358389) (4, 13.200611) (5, 13.251902)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 81.546486) (2, 85.345444) (3, 90.532347) (4, 98.247764) (5, 105.423875)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.539097) (2, 24.432485) (3, 25.629666) (4, 23.935250) (5, 23.911139)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{With precompilation}%
    \label{fig:erase-precomp}
  \end{subfigure}
  \caption{Results of the depth benchmark with precompilation disabled (\subref{fig:erase-noprecomp}) or enabled (\subref{fig:erase-precomp}) and with $a = 0$ (\protect\showpgfsquare) or $a = 100$ (\protect\showpgfcircle).}%
  \label{fig:erase}
\end{figure}
